Reducing Acceptance Marks \\ in Emerson-Lei Automata

### 2021-10-14 paper/main.tex
begining os section 3 modified

\def\Mocc{M_{\mathrm{occur}}}
\def\Mevery{M_{\mathrm{every}}}
\newcommand{\G}{\mathbf{G}}
\newcommand{\CA}{\ensuremath{\mathcal{A}}}
\newcommand{\CB}{\ensuremath{\mathcal{B}}}
\def\init{\iota}
\def\lang{L} % FIX this
\newcommand{\cM}{\mathcal{M}}
\newcommand{\mA}{\mathcal{A}}
\newcommand{\bscc}{\mathcal{B}}
\newcommand{\sched}{\mathfrak{s}}
\newcommand{\Act}{\mathit{Act}}
\newcommand{\NP}{\textsf{NP}}
\newcommand{\POLY}{\textsf{P}}
\newcommand{\class}{\ensuremath{\mathcal{C}}}
\newcommand{\finally}{\mathsf{F}}
\newcommand{\globally}{\mathsf{G}}
% \newcommand{\finally}{\mathbf{F}}
% \newcommand{\globally}{\mathbf{G}}

@@

\subsubsection{Related results}
We are not aware about any existing research focused on simplification
of acceptance formulas of TELA or reduction of the number of its
acceptance marks without increasing the number of states.

@@

\begin{definition}[TELA]
  A \emph{transition-based Emerson-Lei automaton (TELA)} is a tuple
  $\mA =(Q,M,\Sigma,q_I,\delta,\varphi)$, where
  \begin{itemize}
  \item $Q$ is a finite set of \emph{states},
  \item $M$ is a finite set of \emph{acceptance marks},
  \item $\Sigma$ is a finite \emph{alphabet},
  \item $q_I\in Q$ is an \emph{initial state},
  \item $\delta \subseteq Q \times \Sigma \times 2^{M} \times Q$ is a
    \emph{transition relation}, and
  \item $\varphi$ is the \emph{acceptance condition} constructed
    according to the following abstract syntax equation, where $m$
    ranges over $M$:
    $$\varphi ::= \true \mid \false \mid \Inf(m) \mid \Fin(m) \mid \left( \varphi \land \varphi \right) \mid \left( \varphi \lor \varphi \right).$$
\end{itemize}
\end{definition}

%A transition $(p,a,M',q)\in\delta$ is a transition leading from state $p$ to state $q$ labelled with $a$ and acceptance marks $M'$
A run $\pi$ of $\mA$ over an infinite word
$u = u_{0}u_{1}u_{2}\dots \in \Sigma^{\omega}$ is an infinite sequence
of adjacent transitions
$\pi=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots\in\delta^{\omega}$ where
$q_0=q_I$. Let $M(\pi)$ denote set of all acceptance marks that occur
in run $\pi$ in infinitely many transitions. Run $\pi$ is
\emph{accepting} iff it satisfies the formula $\varphi$, where $\pi$
satisfies $\Inf(m)$ iff $m\in M(\pi)$ and $\pi$ satisfies $\Fin(m)$
iff $m\not\in M(\pi)$. The \emph{language} of automaton $\mA$ is the
set $L(\mA)=\{u\in \Sigma^{\omega}\mid\textrm{there is an accepting run of }\mA\textrm{ over }u\}$.
% such that $\mA$ has
% an accepting run over $u$.

@@

For each mark $m\in M$, let $\delta_m$ denotes the set of all
transitions marked with $m$, i.e.,
$\delta_m=\{(p,a,M',q)\in\delta\mid m\in M'\}$. Further, a \emph{path}
from a state $p$ to a state $q$ is a finite sequence of adjacent
transitions of the form
$(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots(q_{n-1},u_{n-1},M_{n-1},q_n)\in\delta^+$
such that $p=q_0$ and $q=q_n$. A nonempty set of states $S\subseteq Q$
is called a nontrivial \emph{strongly connected component (SCC)} if
for each $p,q\in S$ there is a path from $p$ to $q$. An SCC S is
\emph{maximal} if there is no SCC $S'$ satisfying $S\subsetneq S'$. In
the rest of this paper, SCC always refers to a maximal SCC. Given an
SCC $S$, $\delta_S=\delta\cap (S\times \Sigma \times 2^M \times S)$
denotes the set of all transitions between states in $S$.

We often use acceptance marks $\tacc{1}{1},\tacc{2}{2},\ldots\in M$
and we write $\Inf\tacc{4}{j}$ and $\Fin\tacc{8}{k}$ instead of
$\Inf(\tacc{4}{j})$ and $\Fin(\tacc{8}{k})$, respectively. In our
graphical representation of automata,
\begin{tikzpicture}[smallautomaton]
  \node[state] (p) at (0,0) {$p$};
  \node[state] (q) at (1.8,0) {$q$};
  \path [->] (p) edge node[above,pos=.5,overlay] {$a$} node[accset,collacc8, pos=0.65] {k} node[accset,collacc4, pos=0.3] {j} (q);
\end{tikzpicture}
denotes the transition $(p,a,\{\tacc{4}{j},\tacc{8}{k}\},q)\in\delta$.

@@

A QBF $\varphi$ is in the \emph{prenex normal form} if it has the form
$\varphi = Q_{1}x_{1}...Q_{n}x_{n} . \psi$, where
$Q_1,Q_2,\ldots,Q_n \in \{\forall, \exists\}$ and $\psi$ is a Boolean
formula.

@@

In this section, the simplification strategy is to find relations between acceptance sets and reduce the redundant ones. SCC-based simplification is the coarsest simplification we propose.

Let $\mathcal{A}$ be an original automaton and let $\psi$ be an acceptance condition of $\mathcal{A}$. Every run of $\mathcal{A}$ has an infinite suffix that takes place within one SCC $S_{i}$. Thus the evaluation of $\psi$ depends purely on the SCC $S_{i}$ and we can optimize the acceptance condition for each SCC separately. :w
This optimization consists of removing redundant terms from acceptance condition and relabeling of acceptance marks on the transitions. The state and transition structure of the SCC does not change. This way we obtain set of simplified acceptance conditions that we merge into new acceptance formula $\psi'$. The automaton with relabeled transitons and new acceptance condition $\psi'$ we denote $\mathcal{A}'$.  Finally, we ensure that automaton $\mathcal{A'}$ is equivalent to the original automaton $\mathcal{A}$.

\subsection{$S_{i}$ simplification}
\label{section:simplification}
At first, we remove the unused acceptance marks and transform the SCC's acceptance condition $\psi_{i}$ to \emph{disjunctive normal form} (DNF), so we have a unified shape of the condition.
If TELA acceptance condition is in DNF, two distinct terms in one disjunct can only appear in three possible forms:

\begin{itemize}
    \item $\Inf\tacc{2}{i} \land \Inf\tacc{4}{j}$
    \item $\Fin\tacc{8}{i} \land \Fin\tacc{4}{j}$
    \item $\Fin\tacc{8}{i} \land \Inf\tacc{4}{j}$
\end{itemize}

>>>>

\def\Mocc{M_{\mathrm{occur}}}
\def\Mevery{M_{\mathrm{every}}}
\newcommand{\G}{\mathbf{G}}
\newcommand{\CA}{\ensuremath{\mathcal{A}}}
\newcommand{\CB}{\ensuremath{\mathcal{B}}}
\def\init{\iota}
\def\lang{L} % FIX this
\newcommand{\cM}{\mathcal{M}}
\newcommand{\mA}{\mathcal{A}}
\newcommand{\mB}{\mathcal{B}}
\newcommand{\bscc}{\mathcal{B}}
\newcommand{\sched}{\mathfrak{s}}
\newcommand{\Act}{\mathit{Act}}
\newcommand{\NP}{\textsf{NP}}
\newcommand{\POLY}{\textsf{P}}
\newcommand{\class}{\ensuremath{\mathcal{C}}}
\newcommand{\finally}{\mathsf{F}}
\newcommand{\globally}{\mathsf{G}}
% \newcommand{\finally}{\mathbf{F}}
% \newcommand{\globally}{\mathbf{G}}

@@

\subsubsection{Related results}
We are not aware about any existing research aimed at simplification
of acceptance formulas of TELA or reduction of the number of its
acceptance marks without increasing the number of states.

@@

\begin{definition}[TELA]
  A \emph{transition-based Emerson-Lei automaton (TELA)} is a tuple
  $\mA =(Q,M,\Sigma,q_I,\delta,\varphi)$, where
  \begin{itemize}
  \item $Q$ is a finite set of \emph{states},
  \item $M$ is a finite set of \emph{acceptance marks},
  \item $\Sigma$ is a finite \emph{alphabet},
  \item $q_I\in Q$ is an \emph{initial state},
  \item $\delta \subseteq Q \times \Sigma \times 2^{M} \times Q$ is a
    \emph{transition relation}, and
  \item $\varphi$ is the \emph{acceptance condition} constructed
    according to the following abstract syntax equation, where $m$
    ranges over $M$.
    $$\varphi ::= \true \mid \false \mid \Inf(m) \mid \Fin(m) \mid \left( \varphi \land \varphi \right) \mid \left( \varphi \lor \varphi \right)$$
\end{itemize}
\end{definition}

%A transition $(p,a,M',q)\in\delta$ is a transition leading from state $p$ to state $q$ labelled with $a$ and acceptance marks $M'$
A run $\pi$ of $\mA$ over an infinite word
$u = u_{0}u_{1}u_{2}\dots \in \Sigma^{\omega}$ is an infinite sequence
of adjacent transitions
$\pi=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots\in\delta^{\omega}$ where
$q_0=q_I$. Let $M(\pi)$ denote set of all acceptance marks that occur
in run $\pi$ in infinitely many transitions. Run $\pi$ is
\emph{accepting} iff it satisfies the formula $\varphi$, where $\pi$
satisfies $\Inf(m)$ iff $m\in M(\pi)$ and $\pi$ satisfies $\Fin(m)$
iff $m\not\in M(\pi)$. The \emph{language} of automaton $\mA$ is the
set
$L(\mA)=\{u\in \Sigma^{\omega}\mid\textrm{there is an accepting run of
}\mA\textrm{ over }u\}$. Two automata $\mA,\mB$ are \emph{equivalent}
if $L(\mA)=L(\mB)$.
% such that $\mA$ has
% an accepting run over $u$.

@@

A \emph{path} from a state $p$ to a state $q$ is a finite sequence of
adjacent transitions of the form
$(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots(q_{n-1},u_{n-1},M_{n-1},q_n)\in\delta^+$
such that $p=q_0$ and $q=q_n$. A nonempty set of states $S\subseteq Q$
is called a nontrivial \emph{strongly connected component (SCC)} if
for each $p,q\in S$ there is a path from $p$ to $q$. An SCC S is
\emph{maximal} if there is no SCC $S'$ satisfying $S\subsetneq S'$. In
the rest of this paper, SCC always refers to a maximal SCC. Given an
SCC $S$, $\delta_S=\delta\cap (S\times \Sigma \times 2^M \times S)$
denotes the set of all transitions between states in $S$. Further, for
each mark $m\in M$, let $\delta_m$ denotes the set of all transitions
marked with $m$, i.e., $\delta_m=\{(p,a,M',q)\in\delta\mid m\in
M'\}$. The sets $\delta_m$ are called \emph{acceptance sets}.
Finally, given an SCC $S$ and a mark $m\in M$, by $\delta_{S,m}$ we
denote the set $\delta_{S,m}=\delta_S\cap\delta_m$.

We often use acceptance marks $\tacc{1}{1},\tacc{2}{2},\ldots\in M$
and we write $\Inf\tacc{4}{j}$ and $\Fin\tacc{8}{k}$ instead of
$\Inf(\tacc{4}{j})$ and $\Fin(\tacc{8}{k})$, respectively. In our
graphical representation of automata,
\begin{tikzpicture}[smallautomaton,baseline=-2pt]
  \node[state] (p) at (0,0) {$p$};
  \node[state] (q) at (1.8,0) {$q$};
  \path [->] (p) edge node[above,pos=.5,overlay] {$a$} node[accset,collacc8, pos=0.65] {k} node[accset,collacc4, pos=0.3] {j} (q);
\end{tikzpicture}
denotes the transition $(p,a,\{\tacc{4}{j},\tacc{8}{k}\},q)\in\delta$.

@@

A QBF $\varphi$ is in the \emph{prenex normal form} if it has the form
$\varphi = Q_{1}x_{1}...Q_{n}x_{n} . \psi$, where
$Q_1,\ldots,Q_n \in \{\forall, \exists\}$ and $\psi$ is a Boolean
formula.

@@

This section presents a reduction technique based on relations between
acceptance sets restricted to individual SCCs. It employs the fact
that acceptance of a run always depends only on a single SCC, namely
the SCC that contains the whole run except some finite prefix.

Let $\mA$ be a TELA with the acceptance condition $\varphi$. We assume
that $\varphi$ is in DNF. This technique has two steps.
\begin{enumerate}
\item For each SCC $S$, we make a copy $\varphi_S$ of $\varphi$ and
  simplify it according to relations between acceptance sets
  restricted to $\delta_S$. 
\item We collect the simplified acceptance conditions
  $\varphi_{S_1},\ldots,\varphi_{S_k}$ of all SCCs and merge them into
  one acceptance condition $\varphi'$. Additionally, we sometimes add
  acceptance marks to all transitions of some SCCs to make the
  resulting automaton with acceptance formula $\varphi'$ equivalent to
  the original automaton.
\end{enumerate}

% Let $\mathcal{A}$ be an original automaton and let $\psi$ be an
% acceptance condition of $\mathcal{A}$. Every run of $\mathcal{A}$ has
% an infinite suffix that takes place within one SCC $S_{i}$. Thus the
% evaluation of $\psi$ depends purely on the SCC $S_{i}$ and we can
% optimize the acceptance condition for each SCC separately. This
% optimization consists of removing redundant terms from acceptance
% condition and relabeling of acceptance marks on the transitions. The
% state and transition structure of the SCC does not change. This way we
% obtain set of simplified acceptance conditions that we merge into new
% acceptance formula $\psi'$. The automaton with relabeled transitons
% and new acceptance condition $\psi'$ we denote $\mathcal{A}'$.
% Finally, we ensure that automaton $\mathcal{A'}$ is equivalent to the
% original automaton $\mathcal{A}$.

\subsection{Simplification of $\varphi_S$ for an SCC $S$}\label{ssec:step1}
Recall that $\varphi_S$ is in DNF. Hence, we can represent it as a set
$\varphi_S=\{D_1,\ldots,D_k\}$ of disjunctive clauses.
